---
title: "Eyetracking Analysis"
author: "Methods in CCS 25/26"
date: 'Table of Contents:'
output:
  html_document:
    theme: united
    highlight: pygments
    toc: yes
  html_notebook:
    toc: yes
  pdf_document:
    toc: yes
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
runtime: shiny
---

```{=html}
<style>
h1 {
  padding: 25px 0px;
  color: #dd4814;
}
h2 {
  padding: 25px 0px;
}
hr {
  border-color: #dd481433;
  border-style: solid;
  border-width: 4pt 0pt 0pt 0pt;
  margin-top: 20pt;
  margin-bottom: 20pt;
}
.sidestep {
  border: #CCCCCC dotted 3pt;
  border-radius:10pt;
  padding:10pt 10pt;
  margin: 10pt 100pt;
}
accent{
    font-weight: bold;
    font-style: italic;
    text-decoration: underline;
    text-decoration-style:double;
}

</style>
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height = 4)
#NOTE JPB: to knit (while ignoring shiny), just call rmarkdown::render("how_to_analyze_eyetracking_data.rmd")
```

<hr/>

In this document I will demonstrate how to do the basic eye-tracking
analysis steps, focusing on pupil dilation.

To analyze the data from the eye-tracker, we are going to use the
PupilPre package
(<https://cran.r-project.org/web/packages/PupilPre/index.html>). In this
notebook I will demonstrate how to:

-   read in the data
-   remove and interpolate blinks
-   baseline the data
-   make the first plots

PupilPre has more functions, see for example:

-   [Basic Preprocessing of Pupil Size
    Data](https://rdrr.io/cran/PupilPre/f/vignettes/PupilPre_Basic_Preprocessing.Rmd)
-   [Blink and Artifact Cleanup
    (Detailed)](https://rdrr.io/cran/PupilPre/f/vignettes/PupilPre_Cleanup.Rmd)
-   [Interpolation and Filtering of Pupil Size
    Data](https://rdrr.io/cran/PupilPre/f/vignettes/PupilPre_Interpolation_and_Filtering.Rmd)
-   [Aligning Data to a Specific Sample
    Message](https://rdrr.io/cran/PupilPre/f/vignettes/PupilPre_Message_Alignment.Rmd)
-   [Plotting Pupil Size Data Processed with
    PupilPre](https://rdrr.io/cran/PupilPre/f/vignettes/PupilPre_Plotting.Rmd)

### Preparation

First, you need to install the necessary packages:

```{r, eval=FALSE}
install.packages("https://cran.r-project.org/src/contrib/Archive/PupilPre/PupilPre_0.6.2.tar.gz", repos = NULL) #pupilpre no longer active on CRAN, need to install from archive
install.packages("eyelinker")
install.packages("dplyr")
install.packages("plyr")
```

Next, we set the working directory and load the packages:

```{r, eval=TRUE, results='hide', warning=FALSE,message=FALSE}
#set directory
setwd("C:/Users/Gebruiker/Documents/Masters CCS/Year1/Block12/Methods in CCS/algebra/methodsccs-group3/methodsccs-group3/methodsccs_eyetracking/")

#load packages
require(PupilPre)
require(eyelinker)
require(plyr)
require(dplyr)
```

and load the custom `asc_to_pupilpre.R` function that Joshua Krause
provided. This function reads in an .asc file and converts it to the
PupilPre format.

```{r, eval=TRUE}
source("asc_to_pupilpre.R")
```

<hr/>

# Part I: Reading in data

### Step 1. Quick inspection with eyelinker.

First load in the raw data using eyelinker so that it can be inspected.
The raw_dat object contains the raw data as well as information about
whether the eye-tracker detected saccades, blinks, or received messages.
For details the eyelinker vignette can be consulted:
<https://cran.r-project.org/web/packages/eyelinker/vignettes/basics.html>.

```{r, eval=TRUE}
path_to_file <- "sub_1.asc"
raw_dat <- read.asc(path_to_file)
```

Inspect parts of the raw data:

```{r, eval=TRUE}
head(raw_dat$raw)
```

Inspect the messages received by the tracker:

```{r, eval=TRUE}
head(raw_dat$msg, 20)
```

We can infer that the tracker regularly received a "var StartTrial"
message, eventually followed by a "var EndTrial" message.

Eyelinker also parses out fixations, saccades and blinks, e.g.:

```{r, eval=TRUE}
raw_dat$fix
```

If you only want to use fixations, this is the data to use. Note that
you still need to add trial and condition information. Check out all the
way at the bottom of this document on how I did that for the continuous
data. You would need to do something similar here, by first indicating
for each time stamp what condition it belongs to, and then adding that
information to the fixations. Let me know if you need help, or look at
the continuous data as explained below.

Check `str(raw_dat)` to see what other information you get.

<hr/>

### Step 2: preparing the data for the PupilPre package

To get the data into a format that pupilpre can work with, the
asc_to_pupilpre function can be used. This function requires a couple of
parameters:

```{r, eval=TRUE}
label <- "sub1" # Some subject identifier
recorded_eye <- "Left"
trial_start_msg <- "var trial [0-9]+" # Trial begin pattern
id_pattern <- "Start of trial" # Part of trial pattern used for trial index
trial_end_msg <- "var Response given [0-9]+" # Trial end pattern
```

The last three parameters are based on the trial begin and end messages
mentioned earlier. The trial_start_msg should be set to a regular
expression pattern that matches the beginning message.

The function requires a unique numeric identifier in each trial begin
message (just counting total trials works for example). The id_pattern
should match the part of the beginning message just before the number
that identifies the corresponding trial (spaces matter, which is why
id_pattern contains a space right after "StartTrial").

The final trial_end_msg should be set to a regular expression pattern
that matches the end message sent to the tracker after the trial is
over.

Now the function can be used:

```{r, eval=TRUE, message=FALSE}
ppl_pre_dat <- asc_to_pupilpre(path_to_file,
                               label,
                               recorded_eye,
                               trial_start_msg,
                               id_pattern,
                               trial_end_msg)
```

`ppl_pre_dat` contains the pupilpre data, the raw messages, and the raw
data.

```{r, eval=TRUE}
head(ppl_pre_dat$pplpreData)
head(ppl_pre_dat$rawMessages)
head(ppl_pre_dat$rawData)
```

Note that `ppl_prep_data()` is already called within the asc_to_pupilpre
function, so `ppl_pre_dat$pplpreData` can just be used for the remaining
steps in the pupilpre vignettes
(<https://cran.r-project.org/web/packages/PupilPre/index.html>)

<hr/>

### Step 3: extracting experimental data and check messages.

Extract experimental data only:

```{r, eval=TRUE}
dat <- ppl_pre_dat$pplpreData[ppl_pre_dat$pplpreData$in_trial,]
```

Use `str(dat)` to see what info you have.

```{r, eval=TRUE}
str(dat)
```

If you want to see all columns in a Tibble ([a sort of
data.frame](https://r4ds.had.co.nz/tibbles.html)), set the options:

```{r}
options(tibble.width = Inf)
```

Next, we're going to align the data to a certain message in each trial.
Basically, you're indicating what time should be 0 in each trial, which
is typically the stimulus onset (which for this particular dataset was
the end of the fixation cross).

```{r, eval=TRUE}
dat_aligned <- align_msg(dat,Msg="EFixCross")
str(dat_aligned)
```

Note that we now added a column 'Align' with the time relative to the
stimulus.

Next, we create time-series data:

```{r, eval=TRUE, messages=FALSE}
dat_time <- create_time_series(dat_aligned, Adjust = 0)
```

In this case, we didn't need to adjust the timing. However, if your
stimulus starts for example 1000 ms after your aligned message, set
Adjust=1000. See for more info on the above two steps:
<https://rdrr.io/cran/PupilPre/f/vignettes/PupilPre_Message_Alignment.Rmd>.

Check whether the results make sense for our messages:

```{r, eval=TRUE}
#start stimulus
check_msg_time(dat_time, Msg = "EFixCross", ReturnData=T)

#first response
check_msg_time(dat_time, Msg = "MouseClicked")

#varies, so that's good. Should correspond to RT in behavioral data file, and to the RT in the RT_Problem message:
check_msg_time(dat_time, Msg = "RT_Problem")
```

There is certainly a clear relation between the message time of
MouseClicked above and the measured RT_Problem (by OpenSesame), but it's
not exactly the same. This means that somewhere the timing is not
perfect, probably sending the messages to the eyetracker. For
eyetracking this is typically not an issue, but you should be aware of
this if you need specific timing from the eyetracker for, for example,
EEG analysis.

At this point, we're going to indicate that we only want to look at one
eye, the left one if both are recorded:

```{r, eval=TRUE, message=FALSE}
dat2 <- ppl_select_recorded_eye(data = dat_time, Recording = "LorR", WhenLandR = "Left")
```

<hr/>

# Part II: Removing blinks

Now we have aligned the data, we need to clean it. First, we are going
to remove off-screen points: moments when the participant looked next to
the screen.

First plot all fixations, to get an idea of the screen size:

```{r, eval=TRUE, message=FALSE}
with(dat2,plot(Gaze_X,Gaze_Y))
```

The data was probably recorded with a resolution of 1280x1024, remove
anything outside that range:

```{r, eval=TRUE, message=TRUE}
dat3 <- recode_off_screen(data = dat2, ScreenSize = c(1280, 1024))
```

The next question is how many blinks we have:

```{r, eval=TRUE, message=TRUE}
blink_summary(dat3, Summary = "Event")
```

We can also check for missing data in general:

```{r, eval=TRUE, message=TRUE}
NA_summary(dat3, Summary = "Event", PupilColumn="Pupil")
```

This is the same number as above, so all blinks were detected by the SR
Research software.

To remove data associated with blinks, call:

```{r, eval=TRUE, message=FALSE}
dat4 <- clean_blink(dat3, BlinkPadding = c(170, 170), Delta = 5,
                    MaxValueRun = 5, NAsAroundRun = c(2,2),
                    LogFile = "BlinkCleanupLog.rds")
```

This removes blinks, including 170 ms padding on each side. See
<https://rdrr.io/cran/PupilPre/f/vignettes/PupilPre_Cleanup.Rmd>
for details.

Not all blinks and artifacts might be detected by the SR Research
software. If not, you can clean additionally using the clean_artifact
function. Again, see [this
page](hhttps://rdrr.io/cran/PupilPre/f/vignettes/PupilPre_Cleanup.Rmd).

It's very important to not just trust these functions, but check the
results:

```{r, eval=TRUE, message=FALSE}
plot_compare_app(dat4)
```

If not all 'weird' data points are marked in red, increase the padding.
See for example trial `sub3.5`.

Finally, we want to interpolate the now-missing data. For this we can
use linear interpolation or a cubic spline. Let's do the latter:

```{r, eval=TRUE, message=FALSE}
dat5 <- interpolate_NAs(dat4, Method = "linear", XandY = T, MinData = 2)
```

Again, check the results:

```{r, eval=TRUE, message=FALSE}
plot_compare_app(dat5)
```

For details, see [this
page](https://rdrr.io/cran/PupilPre/f/vignettes/PupilPre_Interpolation_and_Filtering.Rmd).

<hr/>

# Part III: Baselining the data

As the next step, we will baseline the data, to remove long term trends
and make the data comparable between subjects.

We use subtractive baselining (see the Mathot 2018 paper), using 500ms
pre-stimulus (the fixation duration):

```{r, eval=TRUE, message=FALSE}
dat6 <- baseline(dat5, BaselineWindow = c(-500, 0), BaselineType = "Subtraction")
```

And finally, we save the data so you don't need to do all these preps
every time:

```{r, eval=FALSE, message=FALSE}
saveRDS(dat6, file = "FinalDat.rds", compress = "xz")
```

<hr/>

# Start plotting!

<https://rdrr.io/cran/PupilPre/f/vignettes/PupilPre_Plotting.Rmd>

For example, the grand average:

```{r, eval=TRUE, message=FALSE}
ppl_plot_avg(data = dat6, xlim = c(-250, 2000), Column = "Pupil",
             Condition1 = NULL, Condition2 = NULL, Cond1Labels = NA, 
             Cond2Labels = NA, ErrorBar = FALSE, ErrorBand=TRUE, PupilPreTheme = TRUE) 
```

Naturally, you can also do this in base plot. First, we calculate the
average:

```{r, eval=TRUE, message=FALSE}
avg <- with(dat6[dat6$Time >= -250 & dat6$Time <= 2000,], aggregate(list(Pupil=Pupil), list(Time=Time),mean))
plot(avg$Time, avg$Pupil)
```

It is also possible to plot by condition. In that case, we first need to
add condition information to our data set:

```{r, eval=TRUE, message=FALSE}
#add condition (this is slow code, you might be able to do better):
dat6$condition <- 0 #ini at 0
cond <- 'none'
for(i in 1:length(dat6$Subject)){ #for each sample in our dataset
    tmp <- dat6$SAMPLE_MESSAGE[i] #check whether there's a message
    if(!is.na(tmp)){ #if there is
        tmp <- regexpr( 'ProblemType [0-9]+', dat6$SAMPLE_MESSAGE[i]) #check whether 'ProblemType ##' is in the message
        if(!is.na(tmp) && tmp != -1){ #if so
            cond <- regmatches(dat6$SAMPLE_MESSAGE[i],tmp)        
            cond <- as.numeric(regmatches(cond,regexpr("[0-9]+",cond))) #set 'cond' to the type 
        }
    }
    dat6$condition[i] <- cond #and add this to the new condition variable that we added above
}
dat6$condition <- as.factor(dat6$condition)

#now we can plot:
ppl_plot_avg(data = dat6, xlim = c(-250, 2000), Column = "Pupil", 
             Condition1 = "condition", Condition2 = NULL, 
             ErrorBar = FALSE, ErrorBand=TRUE, PupilPreTheme = TRUE)
```

And yes, the result isn't pretty based on the one participant in this
data set.

Therefore, the next step is to add more participants. You need to repeat
the preprocessing steps for each participant, and then combine the data
sets after doing the baselining.

Good luck with your analyses!
